\chapter{State of the art}

The standard cosmological model, $\Lambda$CDM, provides a highly successful framework for understanding the large-scale structure of the Universe. However, testing this paradigm on sub-galactic scales reveals persistent theoretical and observational tensions, particularly concerning the internal dark matter distribution of ultra-faint dwarf galaxies. To investigate the formation and evolution of these highly non-linear, dark-matter-dominated systems, cosmologists rely extensively on cosmological $N$-body simulations. The physical fidelity of these computational experiments is fundamentally constrained by numerical choices, most notably the gravitational softening length, which is required to approximate collisionless dynamics and prevent unphysical microscopic particle interactions. This chapter reviews the theoretical and computational foundations necessary to understand the delicate interplay between physical dark matter models and numerical simulation parameters. Section \ref{sec:sota_dynamics} introduces the principles of collisionless galactic dynamics and the formation of dark matter halos. Section \ref{sec:sota_faint_galaxies} outlines the specific small-scale challenges and alternative dark matter paradigms associated with faint galaxies. Section \ref{sec:sota_nbody_softening} details the mechanics of $N$-body simulations and the fundamental necessity of gravitational softening. Section \ref{sec:sota_convergence} explores the historical convergence criteria and the inherent limitations of utilizing a fixed softening length. Section \ref{sec:sota_adaptive_softening} presents the modern evolution towards rigorous, energy-conserving adaptive softening frameworks. Finally, Section \ref{sec:sota_synthesis} synthesizes these concepts to demonstrate how the transition to adaptive softening is critical for accurately resolving the true physical density profiles of ultra-faint dwarf galaxies.

\section{Introduction to collisionless dynamics and dark matter halos}
\label{sec:sota_dynamics}

The study of dark matter halos fundamentally relies on the principles of collisionless galactic dynamics. In the standard cosmological paradigm, cold dark matter behaves as a collisionless fluid where the constituent particles interact almost exclusively through large-scale gravitational forces. A self-gravitating system is considered strictly collisionless when its two-body relaxation time---the timescale over which individual, close-range particle encounters significantly alter a given particle's trajectory---is vastly longer than the dynamical time of the system and the age of the Universe \cite{binney2008galactic}. 

Because discrete, microscopic collisions are negligible in the physical Universe, the evolution of a dark matter halo is governed by the smooth, mean-field gravitational potential generated by the macroscopic mass distribution. The phase-space density of such a collisionless fluid, $f(\mathbf{x}, \mathbf{v}, t)$, is conserved along particle trajectories. This conservation is described mathematically by the collisionless Boltzmann equation (often referred to as the Vlasov equation). To fully determine the system's dynamics, the collisionless Boltzmann equation must be coupled with Poisson's equation, which relates the integrated real-space density to the underlying gravitational potential \cite{binney2008galactic}. 

Through extensive theoretical modeling and early $N$-body simulations, it has been established that dynamically relaxed dark matter halos tend to settle into universal density profiles, regardless of their total mass. These distributions are frequently parameterized by two-power density models \cite{binney2008galactic}, with the most prominent being the Navarro-Frenk-White (NFW) profile. Such models precisely characterize the transition between the shallow inner logarithmic slope and the steeper outer slope of the halo's density distribution. 

Understanding the precise shape of these inner density profiles, specifically whether they exhibit the steep central "cusps" predicted by purely collisionless physics or flattened central "cores", is of great importance. This distinction becomes especially critical, and highly contested, when analyzing the extreme, dark-matter-dominated environments of ultra-faint dwarf galaxies.


\section{The small-scale challenges of faint galaxies}
\label{sec:sota_faint_galaxies}

While the standard Cold Dark Matter ($\Lambda$CDM) paradigm successfully reproduces the large-scale structure of the Universe, it faces persistent observational tensions on sub-galactic scales. Ultra-faint dwarf galaxies, with virial masses on the order of $10^8$ to $10^9 M_{\odot}$, are among the most dark-matter-dominated systems known and thus provide an ideal astrophysical laboratory for testing small-scale structure formation. However, high-resolution cosmological simulations have revealed discrepancies between $\Lambda$CDM predictions and the observed properties of these faint satellites, most notably the apparent deficit of low-mass satellite galaxies, often referred to as the missing satellites problem \cite{salamin2026origin}. Furthermore, the predicted steep central density cusps often conflict with the shallower profiles inferred from kinematic observations, leading to the well-documented core-cusp problem.

To address these small-scale crises, alternative dark matter models have been extensively explored. Warm Dark Matter (WDM) represents a compelling alternative motivated by its potential to naturally reduce the abundance of low-mass halos. In a WDM cosmology, the non-negligible primordial velocities of dark matter particles lead to free-streaming in the early Universe. This free-streaming effect causes a damping of the matter power spectrum at small scales, effectively suppressing the formation of low-mass structures that are ubiquitous in purely cold dark matter models \cite{salamin2026origin}. Consequently, the structural properties of ultra-faint dwarf galaxies are highly sensitive to the assumed nature of the dark matter particle.

Beyond modifications to the fundamental dark matter particle, complex baryonic and dynamical processes can also significantly alter the density profiles and physical sizes of faint galaxies. Recent simulation efforts have suggested that ultra-faint dwarf galaxy analogues formed within the standard $\Lambda$CDM framework can become overly extended relative to the observed luminosity-size relation. One proposed physical mechanism to explain this size growth is dynamical heating, where interactions with numerous low-mass subhalos inject kinetic energy into the central regions of the primary halo \cite{salamin2026origin}. Distinguishing between these genuine physical mechanisms, such as core formation via dynamical heating or free-streaming, and artificial numerical effects in N-body simulations requires immense precision in how the gravitational potential is resolved at the smallest scales.


\section{N-body simulations and the necessity of gravitational softening}
\label{sec:sota_nbody_softening}

To study the nonlinear evolution of dark matter halos, cosmologists rely heavily on N-body simulations. In these numerical experiments, the continuous dark matter phase-space distribution is discretized into a finite number of massive macro-particles. While this approach successfully captures the macroscopic collisionless dynamics described by the collisionless Boltzmann equation, it introduces a fundamental numerical challenge. Because each simulation particle represents an enormous mass of actual dark matter, the simulated system is no longer truly collisionless at microscopic scales.

If raw Newtonian gravity were evaluated between these discrete simulation particles, close encounters would lead to artificial two-body relaxation and unphysical large-angle scattering \cite{binney2008galactic}. These violent numerical collisions would rapidly wash out the delicate dynamical structures of the halo, such as the central density cusp, rendering the simulation entirely unrepresentative of the physical Universe. 

To circumvent this computational artifact, N-body codes introduce the concept of gravitational force softening. By modifying the Newtonian potential at short distances, softening prevents the force between two approaching particles from diverging to infinity. A common approach involves replacing the simple point-mass potential with a smoothed kernel, characterized by a softening length parameter, often denoted as $\epsilon$. This mathematical modification effectively suppresses unphysical singularities and ensures that the simulated system maintains its collisionless nature over the relevant cosmological timescales \cite{power2003inner}. Consequently, the choice of this softening parameter dictates the effective spatial resolution of the simulation, creating a delicate balance between resolving physical small-scale structures and avoiding numerical particle discreteness effects.


\section{Convergence and the limits of fixed softening lengths}
\label{sec:sota_convergence}

Determining the appropriate value for the constant softening length has been the subject of extensive numerical studies. The landmark convergence criteria established by Power et al. \cite{power2003inner} provided a rigorous framework for ensuring that the simulated inner density profiles of dark matter halos are robust and free from numerical artifacts. According to their findings, achieving a convergent mass profile requires a delicate and highly correlated selection of the gravitational softening length, the integration time-step, and the total number of particles. If the chosen softening length is too large, it artificially blurs out physical structures and degrades the spatial resolution of the halo's central region. Conversely, if the softening length is too small, the simulation becomes heavily influenced by artificial two-body relaxation, which exacerbates particle discreteness effects.

While the criteria from Power et al. served as the standard for over a decade, recent high-resolution studies have critically re-evaluated these constraints. Notably, Zhang et al. \cite{zhang2019optimal} argued that the traditional optimal softening scheme is overly conservative for modern cosmological simulations. Through an extensive suite of N-body simulations, they demonstrated that utilizing softening lengths smaller than those recommended by Power et al. can yield numerically convergent results while achieving significantly higher spatial resolution in the density and circular velocity profiles. 

However, Zhang et al. also issued a critical warning regarding the absolute lower limits of this parameter. They discovered that employing an excessively small softening length systematically over-predicts the dark matter density in the innermost regions of the halo. This artificial density enhancement mimics the steep cusps associated with physical cold dark matter, making it highly problematic when attempting to study core formation mechanisms in faint galaxies. Ultimately, this delicate tuning problem highlights the fundamental limitation of employing a single, fixed softening length across a system characterized by the extreme density gradients of a dark matter halo.


\section{The evolution to adaptive gravitational softening}
\label{sec:sota_adaptive_softening}

To overcome the fundamental limitations associated with fixed softening lengths, computational astrophysicists have increasingly turned to adaptive gravitational softening. In systems characterized by vastly different density regimes, such as an ultra-faint dwarf galaxy embedded within a larger cosmological context, a single constant softening length is intrinsically inadequate. Ideally, the softening length should scale dynamically with the local inter-particle spacing, providing high spatial resolution in dense central regions while maintaining sufficient smoothing in the diffuse outskirts to prevent artificial relaxation.

Despite its clear theoretical advantages, the implementation of spatially varying softening lengths historically encountered a severe roadblock. Naive implementations, where each particle is simply assigned a unique softening length based on its local density, fundamentally violate Newton's Third Law of motion. Because the force exerted by a particle on its neighbor would no longer equal the reciprocal force, such simulations suffered from severe non-conservation of both total energy and linear momentum.

A major theoretical breakthrough was achieved by Price and Monaghan \cite{price2007energy}, who successfully derived a consistent adaptive softening formalism from first principles using Lagrangian mechanics. By incorporating the softening length as a strictly defined field variable, their derivation revealed that any spatial variation in the softening length necessitates an additional force term, which is directly proportional to the gradient of the softening length. This extra term acts analogously to pressure forces in Smoothed Particle Hydrodynamics, and it rigorously restores exact energy and momentum conservation within the N-body system.

More recently, this foundational work has been significantly expanded to accommodate the immense complexity of modern, high-resolution cosmological simulations. Hopkins et al. \cite{hopkins2023novel} generalized these conservative methods for multi-species collisionless N-body simulations, which often include distinct interacting components such as dark matter, stars, and black holes. Their state-of-the-art framework ensures consistent interaction terms and timestep criteria across species with different adaptive softening laws, allowing researchers to accurately model the complex, non-linear gravitational interplay in faint galaxies without sacrificing numerical stability or physical conservation laws.


\section{Softening's impact on faint galaxy profiles}
\label{sec:sota_synthesis}

The inner structure of ultra-faint dwarf galaxies is highly sensitive to both the underlying physical paradigm, such as the choice between cold and warm dark matter, and the specific numerical parameters employed in N-body simulations. Because these faint galaxies are characterized by shallow potential wells and relatively low particle counts even in high-resolution zoom-in simulations, they are uniquely vulnerable to the numerical artifacts associated with gravitational softening. 

As demonstrated by Power et al. \cite{power2003inner} and further elucidated by Zhang et al. \cite{zhang2019optimal}, the reliance on a single, fixed softening length inherently compromises the spatial resolution in the innermost regions of the halo or artificially exacerbates two-body relaxation. In the context of ultra-faint dwarf galaxies, an overly conservative fixed softening length can artificially smooth out the central density profile. This numerical blurring can easily mimic the physical core formation predicted by warm dark matter free-streaming or baryonic dynamical heating \cite{salamin2026origin}. Conversely, a softening length that is too small can artificially enhance the central density, falsely reinforcing the steep cusps of the standard cold dark matter model.

Therefore, to confidently isolate and evaluate the true physical mechanisms driving the size growth and density profile evolution of ultra-faint dwarf galaxies, the numerical framework itself must be meticulously controlled. Applying the rigorous, energy-conserving adaptive softening frameworks developed by Price and Monaghan \cite{price2007energy} and Hopkins et al. \cite{hopkins2023novel} is critical to resolving this ambiguity. By utilizing an adaptive gravitational softening length, this thesis aims to ensure that the simulated dark matter halo profiles of faint galaxies are physically accurate and free from the systemic biases of fixed-resolution gravity calculations. Ultimately, bridging this gap between macroscopic cosmological models and the microscopic numerical realities of N-body simulations will provide a highly robust framework for testing the small-scale predictions of modern cosmology.

